---
title: "Package 'MFA'"
author: "Yulin Chen, Stephanie Wuerth, Eren Bilir, Jim Oreluk"
date: "Dec. 3 2016"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
## Overview

Multifactor Analysis is a statistical method that expands the concept of principal component analysis to a multi-table data set. Principal component analysis (PCA) traditionally takes in one 2D data table of observations and variables and uses singular value decomposition (SVD) along with other analytical methods from linear algebra to perform dimension reduction on the data table, or in other words, compute a smaller number of new compound variables called 'components' which can describe the variation seen among the observations. This technique is useful because it gives the user insight into the underlying structure of the data table, or patterns in the data, that would be difficult to see without reducing dimensions. Multifactor Analysis takes this approach further by first doing SVD on each individual table and normalizing each table by its first singular value, which then allows all normalized data tables to be combined into a 'grand' data table which can then itself undergo PCA. A more detailed description of multifactor analysis and a step by step guide to the calculations performed in multifactor analysis can be found in reference 1. 

## Package contents

[The MFA constructor: mfa.R](#id1)

[The print method: print-mfa.R](#id2)

[The four plotting methods: plot_compromise.R, plot_ev.R, plot_loading.R, plot_partial_fac.R](#id3)

[Additional descriptive statistics: lg.R, lg_table.R, rv.R, rv_table.R, obs_dim.R, var_dim.R, bootstrap_factorscores.R](#id4)

[Load a toy data set to experiment with: loadWineData.R](#id5)

[Shiny App: runExample.R](#id6)

[References](#id7)

<a id="id1"></a> 

## The MFA constructor

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

<a id="id2"></a> 

## The print method

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

<a id="id3"></a> 

## Four plotting methods 

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

<a id="id4"></a> 

## Additional descriptive statistics

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

<a id="id5"></a> 

## Toy data set 

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

<a id="id6"></a> 

## Shiny App

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))

<a id="id7"></a> 

## References 

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.
